---
title: "0_data_onboarding"
author: "Rosalee Elting"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is one file that cleans all types of data from 0_data (raw data) and produces products that live in 2_incremental. Those data then could be processed with 1_## code where actual calculations would go.


# LAB FIGHT QUALIFICATION 
Using behavioral coding to characterize tail flaring
##libraries 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(vcd)
library(tibble)
library(stringr)
library(lubridate)
```

##Make a data set 
Qualitative descriptions of Lab trials 
```{r}

fight_qual <- read.csv("0_data/Calliope_Fight_Video_Qual_Descriptions_2023_2025.csv") #this is the full csv from OneDrive, includes data from all 3 years
```
##Process .csv to clean and export for any manual entry 
cleaning function
```{r}
clean_fight_data <- function(df) {
  # Function to extract bird numbers from fight column
  extract_bird_numbers <- function(fight_str) {
    # Remove 'CH' or 'Ch' (case insensitive) and '_r##' or other suffixes
    bird_str <- gsub("(?i)ch|_r\\d+", "", fight_str)
    # Split by underscore and remove any remaining non-numeric characters
    birds <- unlist(strsplit(bird_str, "_"))
    birds <- gsub("[^0-9]", "", birds)
    return(paste(birds, collapse=","))
  }
  
  # Function to get the other bird number
  get_other_bird <- function(bird_nums, retreating_bird) {
    birds <- unlist(strsplit(bird_nums, ","))
    other_bird <- birds[!birds %in% retreating_bird]
    return(other_bird)
  }
  
  # Behavior columns to process
  behavior_cols <- c("Chase", "Signal", "Face.to.face", "Spiral", 
                    "Joust", "Displace", "Retreat")
  
  # Create a copy of the dataframe to modify
  cleaned_df <- df
  
  # Add win column
  cleaned_df$Win <- NA
  
  # Process ALL rows
  for(i in 1:nrow(cleaned_df)) {
    # Extract bird numbers for this fight
    bird_nums <- extract_bird_numbers(cleaned_df$Fight[i])
    
    # Process each behavior column
    for(col in behavior_cols) {
      current_value <- cleaned_df[i, col]
      
      # Replace 'X' with bird numbers
      if(!is.na(current_value) && current_value == "X") {
        cleaned_df[i, col] <- bird_nums
      }
      # Replace 'A' or 'B' with first or second bird number respectively
      else if(!is.na(current_value)) {
        if(current_value == "A") {
          cleaned_df[i, col] <- strsplit(bird_nums, ",")[[1]][1]
        } else if(current_value == "B") {
          cleaned_df[i, col] <- strsplit(bird_nums, ",")[[1]][2]
        }
      }
    }
    
    # Determine winner based on retreat
    retreat_value <- cleaned_df$Retreat[i]
    if(!is.na(retreat_value) && retreat_value != "") {
      # If the retreat value is numeric (a bird number or comma-separated numbers)
      if(grepl("^[0-9,]+$", retreat_value)) {
        # Get the first retreating bird number
        retreating_bird <- strsplit(retreat_value, ",")[[1]][1]
        # The winner is the other bird
        cleaned_df$Win[i] <- get_other_bird(bird_nums, retreating_bird)
      }
    }
  }
  
  return(cleaned_df)
}

# Example usage:
# cleaned_data <- clean_fight_data(fight_qual)
```
## make and export new df
With cleaned data. Can then import in another script after adding any win data manually. 
```{r}
fight_qual_clean <- clean_fight_data(fight_qual)

# select only 2024 and 2025 data
fight_qual_clean <- fight_qual_clean %>%
  filter(year(as.Date(Date, format = "%m/%d/%Y")) >= 2024)

# get a sample size of fights 
fight_qual_clean %>%
  summarise(n_fights = n_distinct(paste(Date, Fight, Run)))

write.csv(fight_qual_clean, "2_incremental/fight_qual_processed.csv")
```

#Arizona TAIL FLARE USAGE
```{r}
az_tail_qual <- read.csv("./0_data/az_tail_spread_qual.csv")
```

##process data 
```{r}
az_tails <- az_tail_qual %>% 
  select(-c(path, X)) %>% 
  filter(!is.na(flaring_present)) %>% #filter out solos
  rename(manuscript_flaring = flaring_present) %>% # renaming column that is the data from the ms, with startle discretion vs. convservative vs liberal startle consideration 
  filter(!is.na(vid_name), vid_name != "") # get rid of rows that are empty and fill in with zeros later. 


az_tails_long <- az_tails %>% # make longer with the qualifier of what conservatisim is coded. 
  pivot_longer(
    cols = c(manuscript_flaring, any_flaring, flare_wo_startle), 
    names_to = "qualifier", 
    values_to = "flare_code"
  )

az_tails_long2 <- az_tails_long %>% 
    mutate( 
    species = species1, #first handle species1 column
    flaring = case_when(
      flare_code == "x" ~ 1, 
      flare_code == "1" ~ 1, 
      flare_code == "2" ~ 0, 
      TRUE ~ 0 #na or blank
    )) %>% 
  select(interaction_id = vid_name, site, species, flaring, qualifier) %>% 
  bind_rows(
    az_tails_long %>% 
      mutate(
        species = species2, 
        flaring = case_when(
          flare_code == "x" ~ 1, 
          flare_code == "1" ~ 0, 
          flare_code == "2" ~ 1,
          TRUE ~ 0
        )
      ) %>% 
      filter(!is.na(vid_name)) %>% #filter out empty rows
      select(interaction_id = vid_name, site, qualifier, species, flaring)
  ) %>%
      arrange(qualifier, interaction_id)

write.csv(az_tails_long2, "./2_incremental/az_tails_long_clean.csv")
```

##df of all AZ Species 
To be used in other code as needed
```{r}
az_species <- tibble(
  species = c("an" ,"bc" ,"bb" ,"bl" ,"ru" ,"ca" ,"ri", "be", "bt"),
  common_name = c("Anna's","Black-chinned", "Broad-billed", "Blue-throated Mountain-gem", "Rufous", "Calliope", "Rivoli's", "Berylline", "Broad-tailed"), 
  sci_name = c("Calypte anna", "Archilochus alexandri", "Cynanthus latirostris", "Lampornis clemenciae", "Selasphorus rufus", "Selasphorus calliope", "Eugenes fulgens", "Saucerottia beryllina", "Selasphorus platycercus")
)

write.csv(az_species, "./2_incremental/az_species_list.csv")
```



#RANKS
```{r}
ranks <- tibble(
  bird_id = c("CH43", "CH44", "CH45", "CH46", "CH47", "CH48", "CH49", "CH50", "CH51", "CH52","CH53","CH54", "CH55", "CH56"),
  dom_rank = c(4, 2, 3, 1, 3, 4, 1, 5, 2, 5, 3,2,1,4)
)
write.csv(ranks, "./2_incremental/bird_ranks.csv")
```

#CALLIOPE TAIL SPREAD ANGLES
```{r}
tail_angles_raw <- read.csv("./0_data/tail_spread_angles_paper_data_2024_2025.csv") # allows for for negative angles from dot products, includes 2024 and 2025 data (n=10)

# handle some uppercase and some lowercase TO/LA between 2024 and 2025
names(tail_angles_raw) <- names(tail_angles_raw) %>% 
  str_replace("_to$", "_TO") %>% 
  str_replace("_la$", "_LA")

# tidy format
# Reshape the data from wide to long format
tail_angles_long <- tail_angles_raw %>%
  rownames_to_column("time_point") %>%
  pivot_longer(
    cols = -time_point,
    names_to = "condition",
    values_to = "angle"
  ) %>%
  # Extract information from the condition column
  mutate(
    time_point = as.numeric(time_point),
    bird_id = str_extract(condition, "ch\\d+"),
    treatment = case_when(
      str_detect(condition, "_TO") ~ "takeoff",
      str_detect(condition, "_LA") ~ "landing",
      str_detect(condition, "_CO") ~ "competition"
    ),
    outcome = case_when(
      str_detect(condition, "_CO_W") ~ "win",
      str_detect(condition, "_CO_L") ~ "loss",
      TRUE ~ NA_character_
    ),
    treatment_outcome = case_when(
    treatment == "competition" & outcome == "win" ~ "Competition (Win)",
    treatment == "competition" & outcome == "loss" ~ "Competition (Loss)",
    treatment == "landing" ~ "Landing",
    treatment == "takeoff" ~ "Takeoff",
    TRUE ~ treatment
  )
  ) 
write.csv(tail_angles_long, "2_incremental/tail_spread_angles.csv")
```

# FRONTAL AREA MORPHOMETRICS
```{r}
bird_area <- read.csv("./0_data/calliope_frontal_area.csv")
```

