---
title: "0_tail_dfs"
author: "Rosalee Elting"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro
This Rmd file is to cleanly add all columns needed to tail angle data from 2024 birds. The input data are from .csvs produced by 0_data_onboarding. This Rmd needs to be run before 1_avg_tail_angles.Rmd. 

# Libraries
```{r, include=FALSE}
library(tidyverse)
library(ggExtra)
library(here)
library(kableExtra)

```

# Input CSV
```{r}
tail_angles_long <- read.csv("2_incremental/tail_spread_angles.csv")

ranks <- read.csv(here("2_incremental", "bird_ranks.csv")) %>% 
  mutate(bird_id = tolower(bird_id)) %>% #make lowercase
  select(-X) #X column not needed

run_data <- read.csv(here("0_data", "tail_paper_video_processing_v2.csv")) # this is the same as the OneDrive tail_paper_video_processing.xls

```

# Create variables used in later mutations 
Max spread angle: assumes some muscular limitation
```{r}
angles <- tail_angles_long %>% 
  filter(!is.na(angle))
max_angle = max(angles$angle)
```

find the shortest run's length (for later subsampling)
```{r}
min_count <- tail_angles_long %>%
  filter(!is.na(angle)) %>%
  count(condition) %>%
  pull(n) %>%
  min()

# counts for all runs 
all_count <- tail_angles_long %>%
  filter(!is.na(angle)) %>%
  count(condition)
```
# format run information
```{r}
run_data <- run_data %>% 
  select(-c(date, digitized, dlc, tail_qc, igor, comments)) %>% 
  mutate(
    bird_id = paste("ch",bird, sep = ""), #bird_id
    run = str_extract(video, "z\\d+$"), #run
    treatment = case_when( 
      str_detect(type, "to") ~ "takeoff",
      str_detect(type, "la") ~ "landing",
      str_detect(type, "co") ~ "competition"), 
    outcome = case_when(
      str_detect(win_lose, "win") ~ "win",
      str_detect(win_lose, "lose") ~ "loss",
      TRUE ~ NA_character_), #outcome and treatment to match tail_angles_long format
    treatment_outcome = case_when(
    treatment == "competition" & outcome == "win" ~ "Competition (Win)",
    treatment == "competition" & outcome == "loss" ~ "Competition (Loss)",
    treatment == "landing" ~ "Landing",
    treatment == "takeoff" ~ "Takeoff",
    TRUE ~ treatment
    ), # outcome to match tail_angles_long format
    condition = case_when(
    treatment == "competition" & outcome == "win" ~ paste0(bird_id, "_", run, "_CO_W"),
    treatment == "competition" & outcome == "loss" ~ paste0(bird_id, "_", run, "_CO_L"),
    treatment == "takeoff" ~ paste0(bird_id, "_", run, "_TO"),
    treatment == "landing" ~ paste0(bird_id, "_", run, "_LA"),
    TRUE ~ paste0(bird_id, "_", run)  # fallback for any other cases
  )
  )%>% 
  filter(
    (bird >= 52) |
      (re.calc_tail_angles == "y")) %>% # only one treatment per bird, from the first RNG version of data, produces 20 entries for 2024. Includes all birds (52-56) from 2025
  select(-c(bird, type, win_lose))
```

# balance data set
Confirm all data (n=40) are frontal approaches

```{r}
frontal <- run_data %>% 
  filter(is.na(rear_approach)) %>% 
  pull(condition)
```

# create table of run variances 
```{r}
run_data_summary <- run_data %>% 
  select(bird_id, treatment_outcome, cv_bill_length, cv_rump_width, pear_cc)

run_data_summary %>%
   kable(digits = 2, 
         col.names = c("Bird ID", "Treatment Outcome", "Bill CV",  "Rump CV", "Rump-Tail Correlation Coefficient"),
        caption = "Summary Statistics for Digitizing Error", 
        format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center",
                font_size = 14) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4472C4")

run_data_summary
save_kable(run_data_summary, "./3_results/digitizing_error_table.jpg")

# write as a csv to use in word
write.csv(run_data_summary, "./2_incremental/digitization_cv_table.csv")

```

# tail_angles_long
All tail angle data for 2024 and 2025 birds, with variable duration of trials
```{r}
tail_angles_long <- tail_angles_long %>%
    filter(bird_id %in% c("ch47", "ch48", "ch49", "ch50", "ch51", "ch52", "ch53", "ch54", "ch55", "ch56")) %>%  #2024 and 2025 birds
    filter(condition %in% frontal) %>% #filter to only frontal approaches
    #filter(condition %in% equal_conditions) %>% #subset for 1 per bird
    mutate(prop_angle = angle/max_angle) %>% 
    left_join(ranks, by = "bird_id") %>% 
    filter(!is.na(angle))   #no flare data, NAs, otherwise all the same length of samples 
```
# redefine the min count with only the 24-25 data 
```{r}
min_count <- tail_angles_long %>%
  filter(!is.na(angle)) %>%
  count(condition) %>%
  pull(n) %>%
  min()
```

# tail_angles_sub
Subset of tail angles so each run is the same length 
```{r,  message = FALSE, echo = FALSE}
set.seed(123)

tail_angles_sub <- tail_angles_long %>% 
  filter(!is.na(angle))  %>%
  group_by(condition) %>% 
  arrange(time_point) %>%
  mutate(interval = floor(n()/ min_count)) %>% #calculates interval to sample
  slice(seq(1, n(), by = first(interval))) %>% #every nth point starting from first
  slice_head(n = min_count) %>% #get exactly min_points, should be very close already
  mutate(time_point_sub = row_number()/min_count) %>% #row number for subset, goes from 0 to 1, a fraction of way to min count.
  ungroup()
# check that all conditions ahve the same nubmer of rows (equals min_count)
tail_angles_sub %>%
  count(condition) %>% 
  print(n=40)
```

# bird_treatment_angle
Summary of mean and max tail spread angle for each bird in each trial. From full, not subset, data.
```{r}
bird_treatment_angle <- tail_angles_long %>%
  group_by(bird_id, treatment_outcome) %>% #not condition bc may want to increase n 
  summarize(mean_angle = mean(angle, na.rm = TRUE),
            max_angle = max(angle, na.rm = TRUE),
            sd_angle = sd(angle, na.rm = TRUE),
            dom_rank = max(dom_rank, na.rm = TRUE),
            .groups = "drop")
```


# Export all dfs 
```{r}
write.csv(tail_angles_long, "./2_incremental/tail_angles_long.csv")
write.csv(tail_angles_sub, "./2_incremental/tail_angles_sub.csv")
write.csv(bird_treatment_angle , "./2_incremental/bird_treatment_angle.csv")
```

