---
title: "1_tail_angles_viz"
author: "Rosalee Elting"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Intro 
For 2024 and 2025 birds, with a representative of a win, loss, takeoff and landing from each individual 
with data frames prepared in 0_tail_dfs.Rmd 

# Libraries
```{r, include=FALSE}
library(tidyverse)
library(ggExtra)
library(here)
library(knitr)
library(pracma)
library(MetBrewer)
library(zoo)
library(ggsignif)
library(ggnewscale)
```

# Load Data 
load data if not already in environment. 
Need data further processed form 0_tail_dfs. 

```{r, include = FALSE}
tail_angles_long <- read.csv("C:/Users/rosal/Documents/R/ProcB_TailFlaring/2_incremental/tail_angles_long.csv")
tail_angles_sub <- read.csv("C:/Users/rosal/Documents/R/ProcB_TailFlaring/2_incremental/tail_angles_sub.csv")
bird_treatment_angle<- read.csv("C:/Users/rosal/Documents/R/ProcB_TailFlaring/2_incremental/bird_treatment_angle.csv")

```
# Color Selection
```{r}
cat_col <- met.brewer("Nattier")[c(2,3, 5, 7)]
bird_col <- met.brewer("Juarez", n=10, direction = -1) # formerly Egypt but not enough contrast. 
contrast_col <- met.brewer("Paquin")[c(3,9)]

# Preview
par(mfrow=c(3,1))
scales::show_col(cat_col, labels=FALSE)
scales::show_col(bird_col, labels=FALSE)
scales::show_col(contrast_col, labels=FALSE)
```

# Fig 2: Mean tail spread by 
Bird's data pooled. As a box plot of means to then add statistical differences in between 
Stats based on n=10, mean from each individual per condition. 

```{r, message = FALSE, echo = FALSE}
# Bracket 1: Take off vs. pooled competition
bracket1_y <- 170      # height of the horizontal bar
bracket1_drop <- 15   # how far it drops to takeoff box 
tip1 <- 2          # small tick at right end

# Bracket 2: win vs lose. 
bracket2_y <- 155       # horizontal bar of these two
tip2 <- 4     # tick at each end 
label2_offset <-2 # between bar and label 

# manage palette
bird_col <- as.character(bird_col)
names(bird_col) <- paste("Bird", 1:10)

# split into years
bird_col_y1 <- bird_col[paste("Bird", 1:5)]
bird_col_y2 <- bird_col[paste("Bird", 6:10)]

# bird id based on bird 1-10 not ch numbers 
bird_remap <- c(
  "ch47" = "Bird 1", "ch48" = "Bird 2", "ch49" = "Bird 3",
  "ch50" = "Bird 4", "ch51" = "Bird 5", "ch52" = "Bird 6",
  "ch53" = "Bird 7", "ch54" = "Bird 8", "ch55" = "Bird 9", "ch56" = "Bird 10"
)

# year shapes
year_shapes <- c(
  "Bird 1" = 19, "Bird 2" = 19, "Bird 3" = 19, "Bird 4" = 19, "Bird 5" = 19,
  "Bird 6" = 15,  "Bird 7" = 15,  "Bird 8" = 15,  "Bird 9" = 15,  "Bird 10" = 15
)
 # update data frames 
# Remap bird_id in both dataframes
tail_angles_sub <- tail_angles_sub %>%
  mutate(bird_id = recode(bird_id, !!!bird_remap),
         bird_id = factor(bird_id, levels = paste("Bird", 1:10)),
         year = ifelse(bird_id %in% paste("Bird", 1:5), "Year 1", "Year 2"))

bird_treatment_angle <- bird_treatment_angle %>%
  mutate(bird_id = recode(bird_id, !!!bird_remap),
         bird_id = factor(bird_id, levels = paste("Bird", 1:10)),
         year = ifelse(bird_id %in% paste("Bird", 1:5), "Year 1", "Year 2"))

bird_treatment_angle %>% # summary from just the n=10 means 
  ggplot(aes(x= treatment_outcome, y = mean_angle))+ 
  geom_boxplot(alpha = 0.3, fill="gray80", color= "gray40")+
  
  # Year 1 points
  geom_pointrange(data = filter(bird_treatment_angle, year == "Year 1"),
                  aes(x = treatment_outcome, y = mean_angle,
                      ymin = mean_angle - sd_angle,
                      ymax = mean_angle + sd_angle,
                      color = bird_id, shape = bird_id),
                  position = position_jitter(width = 0.15, seed = 123),
                  size = 0.6, linewidth = 0.7) +
  scale_color_manual(name = "Year 1", values = bird_col_y1,
                     labels = paste("Bird", 1:5)) +
  scale_shape_manual(name = "Year 1", values = setNames(rep(19, 5), paste("Bird", 1:5)),
                     labels = paste("Bird", 1:5)) +
  
  new_scale_color() +
  new_scale("shape") +
  
  # Year 2 points (open circles)
  geom_pointrange(data = filter(bird_treatment_angle, year == "Year 2"),
                  aes(x = treatment_outcome, y = mean_angle,
                      ymin = mean_angle - sd_angle,
                      ymax = mean_angle + sd_angle,
                      color = bird_id, shape = bird_id),
                  position = position_jitter(width = 0.15, seed = 123),
                  size = 0.6, linewidth = 0.7) +
  scale_color_manual(name = "Year 2", values = bird_col_y2,
                     labels = paste("Bird", 6:10)) +
  scale_shape_manual(name = "Year 2", values = setNames(rep(15, 5), paste("Bird", 6:10)),
                     labels = paste("Bird", 6:10)) +

  scale_x_discrete(limits = c("Takeoff", "Landing", "Competition (Loss)", "Competition (Win)"), 
                   labels = c("Competition (Loss)" = "Competition\n(Loss)",
                              "Competition (Win)" = "Competition\n(Win)",
                              "Landing" = "Landing",
                              "Takeoff" = "Takeoff"))+ 

    # bracket 1
  annotate("segment", x = 1, xend = 3.5, y = bracket1_y, yend = bracket1_y, color = "black") +
  annotate("segment", x = 1, xend = 1, y = bracket1_y - bracket1_drop, yend = bracket1_y, color = "black") +
  annotate("segment", x = 3.5, xend = 3.5, y = bracket2_y + 8, yend = bracket1_y, color = "black") +
  annotate("text", x = 2.25, y = bracket1_y + 4, label = "***", color = "black", size = 6) +

  # bracket 2 
  annotate("segment", x = 3, xend = 4, y = bracket2_y, yend = bracket2_y, color = "black") +
  annotate("segment", x = 3, xend = 3, y = bracket2_y - tip2, yend = bracket2_y, color = "black") +
  annotate("segment", x = 4, xend = 4, y = bracket2_y - tip2, yend = bracket2_y, color = "black") +
  annotate("text", x = 3.5, y = bracket2_y + label2_offset, label = "**", color = "black", size = 6) +
  theme_minimal() +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 16), 
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14), 
    legend.title = element_text(size = 20)
  ) + 
  labs(y = "Tail Angle(\u00B0)",
       x = NULL)
```
# Figure 3: Frequency distibution of tail flare angles
```{r}
tail_angles_sub %>% 
  mutate(treatment_outcome = factor(treatment_outcome, 
                                    levels = c("Competition (Win)", "Competition (Loss)", "Takeoff", "Landing")),
         year = ifelse(bird_id %in% paste("Bird", 1:5), "Year 1", "Year 2")) %>%
  ggplot(aes(x = angle)) +
  
  # Individual bird areas
  # Year 1
  stat_bin(data = . %>% filter(year == "Year 1"),
           aes(color = bird_id, fill = bird_id, y = after_stat(count)),
           binwidth = 5, geom = "line", alpha = 0.5, linewidth = 0.75, position = "identity") + # switch between geom=line and geom=area, will adjust alpha value also 
  scale_color_manual(name = "Year 1", values = bird_col_y1, labels = paste("Bird", 1:5)) +
  scale_fill_manual(name = "Year 1", values = bird_col_y1, labels = paste("Bird", 1:5)) +
  
  new_scale_color() +
  new_scale_fill() +
  
  # Year 2
  stat_bin(data = . %>% filter(year == "Year 2"),
           aes(color = bird_id, fill = bird_id, y = after_stat(count)),
           binwidth = 5, geom = "line", alpha = 0.5, linewidth = 0.75, position = "identity") +
  scale_color_manual(name = "Year 2", values = bird_col_y2, labels = paste("Bird", 6:10)) +
  scale_fill_manual(name = "Year 2", values = bird_col_y2, labels = paste("Bird", 6:10)) +
  
  # Black summary line for all birds combined
  geom_freqpoly(binwidth = 5, color = "black", linewidth = 0.8) +
  
  theme_minimal() +
  labs(x = "Tail Angle(\u00B0)", y = "Count") +
  facet_wrap(~ treatment_outcome, ncol = 1, scales = "free_y") +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 16),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16)
  )
```

# Fig 4: Call and Response of flaring
a single run of the same length: ch47_ch49_z17
```{r}
fig_sm <-  tail_angles_long %>% 
  filter(time_point <= 1190) %>%
  mutate(time_point_s = time_point/2000) %>%
  filter(condition %in% c("ch47_z17_CO_L", "ch49_z17_CO_W")) %>%
  arrange(bird_id, time_point_s) %>%
  group_by(bird_id) %>%
  mutate(angle_smooth = rollmean(angle, k = 5, fill = NA, align = "center")) %>%
  ungroup()

# points I would like to highlight in winner and loser
loser_highlight <- c(0.1125, 0.280, 0.400, 0.595)
winner_highlight <- c(0.0630, 0.235)

loser_highlight_data <- fig_sm %>% 
  filter(bird_id == "ch47", 
    time_point_s %in% loser_highlight)

winner_highlight_data <- fig_sm %>% 
  filter(bird_id == "ch49", 
    time_point_s %in% winner_highlight)
```

```{r}
fig_sm %>% 
  ggplot(aes(x = (time_point_s*1000), y = angle_smooth, color = bird_id))+
  geom_line(linewidth = 0.5, alpha =0.6)+ #smoothed data, light to catch tail pumping
  geom_smooth(method =  "loess", span = 0.15, se = FALSE, linewidth = 1.5) + #trend line
  scale_color_manual(values = bird_col, labels = c("ch47" = "Loser", 
                             "ch49" = "Winner"))+
  labs(x = "Time (ms)",
       y = "Tail Angle (\u00B0)",
       color = "Competitor")+ 
  theme_minimal()+
  theme(
    legend.position = "right",
    strip.text = element_text(size = 24, face = "bold"),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24), 
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 14), 
    legend.title = element_text(size = 20)
  )

```
