---
title: "4_fight_qual_stats"
author: "Rosalee Elting"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r}
library(tidyverse)
library(PMCMRplus)
```

# Import data 
```{r}
fights_long <- read.csv("C:/Users/rosal/Documents/R/ProcB_TailFlaring/2_incremental/fights_long.csv")
```

# Tables of relationship between tail flare and approach angle. 
table of approach proportions instead of counts to control for pseudoreplication
```{r}
individuals <- c("CH47", "CH48", "CH49", "CH50", "CH51", "CH52", "CH53", "CH54", "CH55", "CH56")

fight_summary <- fights_long %>%
  filter(bird_id %in% individuals) %>% 
  group_by(bird_id) %>%
  summarise(
    prop_signal_rear    = sum(Signal[Rear.approach == 1], na.rm = TRUE) / sum(Rear.approach == 1, na.rm = TRUE),
    prop_signal_side    = sum(Signal[Side.approach == 1], na.rm = TRUE) / sum(Side.approach == 1, na.rm = TRUE),
    prop_signal_frontal = sum(Signal[Frontal.approach == 1], na.rm = TRUE) / sum(Frontal.approach == 1, na.rm = TRUE)
  )
```

Pivoting table longer
```{r}
fight_summary_long <- fight_summary %>%
  pivot_longer(cols = c(prop_signal_rear, prop_signal_side, prop_signal_frontal),
               names_to = "approach",
               values_to = "prop_signaling") %>%
  filter(!is.nan(prop_signaling)) %>% 
  filter(!bird_id %in% c("CH50", "CH52")) # ch50 and ch52 do not have samples of rear approach==1, so there is not a representation of all approach angles, let alone if signaling was/wasn't used before. 


```
now three columns for each bird x approach angle combination 

# Statistical analysis of signaling proportion and approach angle 
## Friedman test 
Association between tail angle and approach angle. 
```{r}
friedman.test(prop_signaling ~ approach | bird_id, data = fight_summary_long)
```
## Conover test 
Which relationships drive this Friedman test result 
```{r}
frdAllPairsConoverTest(y = fight_summary_long$prop_signaling, 
                       groups = fight_summary_long$approach, 
                       blocks = fight_summary_long$bird_id,
                       p.adjust.method = "BH")

```

# Statistical analysis of approach angle and win likelihood 
First, wrangle data 
```{r}
individuals <- c("CH47", "CH48", "CH49", "CH50", "CH51", "CH52", "CH53", "CH54", "CH55", "CH56")

fight_summary_2 <- fights_long %>%
  filter(bird_id %in% individuals) %>% 
  group_by(bird_id) %>%
  summarise(
    prop_win_rear    = sum(Win[Rear.approach == 1], na.rm = TRUE) / sum(Rear.approach == 1, na.rm = TRUE),
    prop_win_side    = sum(Win[Side.approach == 1], na.rm = TRUE) / sum(Side.approach == 1, na.rm = TRUE),
    prop_win_frontal = sum(Win[Frontal.approach == 1], na.rm = TRUE) / sum(Frontal.approach == 1, na.rm = TRUE)
  )
```

```{r}
fight_summary_long_2 <- fight_summary_2 %>%
  pivot_longer(cols = c(prop_win_rear, prop_win_side, prop_win_frontal),
               names_to = "approach",
               values_to = "prop_win") %>% 
  filter(!is.nan(prop_win)) %>% 
  filter(!bird_id %in% c("CH50", "CH52")) # ch50 and ch52 do not have samples of rear approach==1, let alone the win combination. Omitting these birds for this test also. 
```
## Friedman Test

```{r}
friedman.test(prop_win ~ approach | bird_id, data = fight_summary_long_2)
```
