---
title: "1_frontal_area_report"
author: "Rosalee Elting"
date: "2025-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries 
```{r}
library(dplyr)
library(tidyverse)
library(lme4)
library(rstatix)
```

# load data 
```{r}
body_mass_adj <- read.csv("./0_data/body_mass_adj.csv")
```
bird_area df should be loaded first from 0_data_onboarding code


# Calculate frontal areas and proportions
```{r}
bird_area_props <- bird_area %>% 
  rename(bird_id = X) %>% 
  select(-comments) %>% 
  mutate(area_relax = body_area_wo_tail + tail_area_relax, 
         area_flare = body_area_wo_tail + tail_area_flared, 
         inc_size = (area_flare/area_relax)-1)
```

# Reporting averages
```{r}
mean(bird_area_props$area_relax)
sd(bird_area_props$area_relax)
mean(bird_area_props$area_flare)
sd(bird_area_props$area_flare)
mean(bird_area_props$inc_size)
sd(bird_area_props$inc_size)
```

# Stats to see if flare state increases perceived size 
So, using a Wilcoxson signed rank test (non-parametric equivalent of a t-test)
testing difference in body area with relaxed vs flared tail 
```{r}
relax <- bird_area_props$area_relax
flare <- bird_area_props$area_flare

wilcox.test(relax, flare, paired = TRUE, alternative = "less") # expect the data to be less in relaxed than flared
```
p-value = 0.0009766

Calculate effect size 
```{r}
# combine into a data frame first
df <- data.frame(
  value     = c(relax, flare),
  condition = rep(c("relax", "flare"), each = 10),
  bird      = rep(1:10, 2)
)
wilcox_effsize(df, value ~ condition, paired = TRUE)
```
Effect size, r= 0.886, large effect size. 

# Stats to see if increase in perceived size due to body mass
This would be an estimate if this is an "index signal" 
These data are from 1_body_mass_RHP.Rmd file 
The data frame is body_mass_adj, which is a series of log-transformed body masses for the whole summer's worth of measures. 

```{r}
individuals <- c("CH47", "CH48", "CH49", "CH50", "CH51", "CH52", "CH53", "CH54", "CH55", "CH56")

mass_summary <- body_mass_adj %>% 
  filter(bird_id %in% individuals) %>% 
  filter(fight_day == "y") %>% # don't need W/L even, but want to use a time frame where fights were occurring. 
  group_by(bird_id) %>% 
  reframe(mean_mass = mean(mass),
          sd_mass = sd(mass))
```
Mass summary to report in paper 
```{r}
paper_mass <- body_mass_adj %>% 
  filter(bird_id %in% individuals) %>% 
  filter(fight_day == "y")

```
Test to see if the tail size is correlated with body size. 
Using flared tail area to test prediction 
```{r}
# merging df to ensure the order of bird_id is the same for paired test 
flare_mass_data <- left_join(
  mutate(bird_area_props, join_key = tolower(bird_id)),
  mutate(mass_summary, join_key = tolower(bird_id)),
  by = "join_key"
) %>%
  select(-join_key) %>% 
  mutate(bird_id = bird_id.x)

# paired data now in vectors 
flare <- flare_mass_data$tail_area_flared
mass <- flare_mass_data$mean_mass
```

```{r}
cor.test(mass, flare, method = "spearman")
```
rho = -0.2, p-value = 0.5835


