---
title: "1_tail_angles_stats"
author: "Rosalee Elting"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Read Data 
```{r}
tail_angles_sub <- read.csv("./2_incremental/tail_angles_sub.csv")
tail_angles_long <- read.csv("./2_incremental/tail_angles_long.csv")
```

# Models
```{r}
library(dplyr)
library(lme4)
library(tidyverse)
library(plotrix)
library(rstatix)
```


# Updated Models 
Starting by understanding the distribution of all the data, as well as the subset data

```{r}
ggplot(tail_angles_long, aes(x = angle)) + 
  geom_histogram()

ggplot(tail_angles_sub, aes(x = angle)) + 
  geom_histogram()
```
Both data sets have a left-skewed data distribution that goes on both sides of zero (many tail angles are sub zero)

T-tests of any form should be normally distributed, but with large sample sizes they are often robust enough. Starting with some t-tests. 


# Test 1: Wilcoxon's Signed Rank Paired t-test
To use an individual bird's mean tail angle for a takeoff compared to a competition 
Making data frames for each bird's take off and landing score. 

## Pooled: TO+LA vs Comp 
```{r}
solitary <- c("takeoff", "landing")

# bird 49
  ch49_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_so <- ch49_so[[1]]
  
  ch49_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_co <- ch49_co[[1]]
  b1_so <- mean(ch49_so)
  b1_co <- mean(ch49_co)

# bird 51
  ch51_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_so <- ch51_so[[1]]
  
  ch51_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_co <- ch51_co[[1]]
  b2_so <- mean(ch51_so)
  b2_co <- mean(ch51_co)
  
# bird 47
  ch47_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_so <- ch47_so[[1]]
  
  ch47_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_co <- ch47_co[[1]]
  b3_so <- mean(ch47_so)
  b3_co <- mean(ch47_co)
  
# bird 48
  ch48_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_so <- ch48_so[[1]]
  
  ch48_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_co <- ch48_co[[1]]
  b4_so <- mean(ch48_so)
  b4_co <- mean(ch48_co)
  
# bird 50 
  ch50_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_so <- ch50_so[[1]]
  
  ch50_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_co <- ch50_co[[1]]
  b5_so <- mean(ch50_so)
  b5_co <- mean(ch50_co)
  
# bird 55
  ch55_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_so <- ch55_so[[1]]
  
  ch55_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_co <- ch55_co[[1]]
  b6_so <- mean(ch55_so)
  b6_co <- mean(ch55_co)
  
# bird 54
  ch54_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_so <- ch54_so[[1]]
  
  ch54_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_co <- ch54_co[[1]]
  b7_so <- mean(ch54_so)
  b7_co <- mean(ch54_co)

# bird 53
  ch53_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_so <- ch53_so[[1]]
  
  ch53_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_co <- ch53_co[[1]]
  b8_so <- mean(ch53_so)
  b8_co <- mean(ch53_co)
  
# bird 56
  ch56_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_so <- ch56_so[[1]]
  
  ch56_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_co <- ch56_co[[1]]
  b9_so <- mean(ch56_so)
  b9_co <- mean(ch56_co)
  
# bird 52
  ch52_so <- tail_angles_sub %>% 
    filter(treatment %in% solitary) %>% 
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_so <- ch52_so[[1]]
  
  ch52_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_co <- ch52_co[[1]]
  b10_so <- mean(ch52_so)
  b10_co <- mean(ch52_co)
  
  

so <- c(b1_so, b2_so, b3_so, b4_so, b5_so, b6_so, b7_so, b8_so, b9_so, b10_so) 
co <- c(b1_co, b2_co, b3_co, b4_co, b5_co, b6_co, b7_co, b8_co, b9_co, b10_co)
```
Testing for differences in the mean for each bird 
```{r}
wilcox.test(so, co, paired = TRUE)
```

# Test 2: Takeoff versus competition 
```{r}
# bird 49
  ch49_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_to <- ch49_to[[1]]
  
  ch49_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_co <- ch49_co[[1]]
  b1_to <- mean(ch49_to)
  b1_co <- mean(ch49_co)

# bird 51
  ch51_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_to <- ch51_to[[1]]
  
  ch51_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_co <- ch51_co[[1]]
  b2_to <- mean(ch51_to)
  b2_co <- mean(ch51_co)
# bird 47
  ch47_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_to <- ch47_to[[1]]
  
  ch47_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_co <- ch47_co[[1]]
  b3_to <- mean(ch47_to)
  b3_co <- mean(ch47_co)
# bird 48
  ch48_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_to <- ch48_to[[1]]
  
  ch48_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_co <- ch48_co[[1]]
  b4_to <- mean(ch48_to)
  b4_co <- mean(ch48_co)
# bird 50 
  ch50_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_to <- ch50_to[[1]]
  
  ch50_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_co <- ch50_co[[1]]
  b5_to <- mean(ch50_to)
  b5_co <- mean(ch50_co)

  # bird 55
  ch55_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_to <- ch55_to[[1]]
  
  ch55_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_co <- ch55_co[[1]]
  b6_to <- mean(ch55_to)
  b6_co <- mean(ch55_co)

# bird 54
  ch54_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_to <- ch54_to[[1]]
  
  ch54_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_co <- ch54_co[[1]]
  b7_to <- mean(ch54_to)
  b7_co <- mean(ch54_co)

# bird 53
  ch53_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_to <- ch53_to[[1]]
  
  ch53_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_co <- ch53_co[[1]]
  b8_to <- mean(ch53_to)
  b8_co <- mean(ch53_co)

# bird 56
  ch56_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_to <- ch56_to[[1]]
  
  ch56_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_co <- ch56_co[[1]]
  b9_to <- mean(ch56_to)
  b9_co <- mean(ch56_co)

# bird 52
  ch52_to <- tail_angles_sub %>% 
    filter(treatment == "takeoff") %>% 
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_to <- ch52_to[[1]]
  
  ch52_co <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_co <- ch52_co[[1]]
  b10_to <- mean(ch52_to)
  b10_co <- mean(ch52_co)

to <- c(b1_to, b2_to, b3_to, b4_to, b5_to, b6_to, b7_to, b8_to, b9_to, b10_to) 
co <- c(b1_co, b2_co, b3_co, b4_co, b5_co, b6_co, b7_co, b8_co, b9_co, b10_co)
```


Testing for differences in the mean for each bird 
```{r}
wilcox.test(to, co, paired = TRUE, alternative = "less")
```
This tests the directional hypothesis that our a priori expectation would be for competition to be greater than landing. 

## Test 2 effect size
I want to obtain an effect size, so using an alternative R package for the same test. 
```{r}
# combine into a data frame first
df <- data.frame(
  value     = c(to, co),
  condition = rep(c("takeoff", "competition"), each = 10),
  bird      = rep(1:10, 2)
)
wilcox_effsize(df, value ~ condition, paired = TRUE)
```
This is a large effect size of 0.886, so not only SS but a large magnitude difference using rank-biserial correlation

# Test 3: Takeoff versus landing 

```{r}
# bird 49
  ch49_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_la <- ch49_la[[1]]
b1_la <- mean(ch49_la)


# bird 51
  ch51_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_la <- ch51_la[[1]]
b2_la <- mean(ch51_la)

# bird 47
  ch47_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_la <- ch47_la[[1]]
b3_la <- mean(ch47_la)

# bird 48
  ch48_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_la <- ch48_la[[1]]
b4_la <- mean(ch48_la)

# bird 50 
  ch50_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_la <- ch50_la[[1]]
b5_la <- mean(ch50_la)

# bird 55
  ch55_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_la <- ch55_la[[1]]
b6_la <- mean(ch55_la)

# bird 54
  ch54_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_la <- ch54_la[[1]]
b7_la <- mean(ch54_la)

# bird 53
  ch53_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_la <- ch53_la[[1]]
b8_la <- mean(ch53_la)

# bird 56
  ch56_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_la <- ch56_la[[1]]
b9_la <- mean(ch56_la)

# bird 52
  ch52_la <- tail_angles_sub %>% 
    filter(treatment == "landing") %>% 
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_la <- ch52_la[[1]]
b10_la <- mean(ch52_la)

la <- c(b1_la, b2_la, b3_la, b4_la, b5_la, b6_la, b7_la, b8_la, b9_la, b10_la)

```

```{r}
wilcox.test(to, la, paired = TRUE)
```
## Test 3 effect size

```{r}
# combine into a data frame first
df <- data.frame(
  value     = c(to, la),
  condition = rep(c("takeoff", "landing"), each = 10),
  bird      = rep(1:10, 2)
)
wilcox_effsize(df, value ~ condition, paired = TRUE)
```
The effect size here is quite small, 0.0483. 

# Test 4: winning versus losing 

```{r}
# bird 49
ch49_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch49") %>% 
    select(angle)
  ch49_l <- ch49_l[[1]]
  
ch49_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch49") %>% 
    select(angle)
ch49_w <- ch49_w[[1]]
  b1_w <- mean(ch49_w)
  b1_l <- mean(ch49_l)

# bird 51
ch51_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch51") %>% 
    select(angle)
  ch51_l <- ch51_l[[1]]
  
ch51_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch51") %>% 
    select(angle)
ch51_w <- ch51_w[[1]]

b2_w <- mean(ch51_w)
b2_l <- mean(ch51_l)
  
# bird 47
ch47_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch47") %>% 
    select(angle)
  ch47_l <- ch47_l[[1]]
  
ch47_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch47") %>% 
    select(angle)
ch47_w <- ch47_w[[1]]

b3_w <- mean(ch47_w)
b3_l <- mean(ch47_l)
# bird 48
  ch48_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch48") %>% 
    select(angle)
  ch48_l <- ch48_l[[1]]
  
ch48_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch48") %>% 
    select(angle)
ch48_w <- ch48_w[[1]]

b4_w <- mean(ch48_w)
b4_l <- mean(ch48_l)
# bird 50 
  ch50_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch50") %>% 
    select(angle)
  ch50_l <- ch50_l[[1]]
  
ch50_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch50") %>% 
    select(angle)
ch50_w <- ch50_w[[1]]

b5_w <- mean(ch50_w)
b5_l <- mean(ch50_l)

# bird 55
ch55_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch55") %>% 
    select(angle)
  ch55_l <- ch55_l[[1]]
  
ch55_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch55") %>% 
    select(angle)
ch55_w <- ch55_w[[1]]
b6_w <- mean(ch55_w)
b6_l <- mean(ch55_l)

# bird 54
ch54_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch54") %>% 
    select(angle)
  ch54_l <- ch54_l[[1]]
  
ch54_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch54") %>% 
    select(angle)
ch54_w <- ch54_w[[1]]
b7_w <- mean(ch54_w)
b7_l <- mean(ch54_l)

# bird 53
ch53_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch53") %>% 
    select(angle)
  ch53_l <- ch53_l[[1]]
  
ch53_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch53") %>% 
    select(angle)
ch53_w <- ch53_w[[1]]
b8_w <- mean(ch53_w)
b8_l <- mean(ch53_l)

# bird 56
ch56_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch56") %>% 
    select(angle)
  ch56_l <- ch56_l[[1]]
  
ch56_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch56") %>% 
    select(angle)
ch56_w <- ch56_w[[1]]
b9_w <- mean(ch56_w)
b9_l <- mean(ch56_l)

# bird 52
ch52_l <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "loss") %>%
    filter(bird_id == "ch52") %>% 
    select(angle)
  ch52_l <- ch52_l[[1]]
  
ch52_w <- tail_angles_sub %>% 
    filter(treatment == "competition") %>% 
    filter(outcome == "win") %>%
    filter(bird_id == "ch52") %>% 
    select(angle)
ch52_w <- ch52_w[[1]]
b10_w <- mean(ch52_w)
b10_l <- mean(ch52_l)

w <- c(b1_w, b2_w, b3_w, b4_w, b5_w, b6_w, b7_w, b8_w, b9_w, b10_w) 
l <- c(b1_l, b2_l, b3_l, b4_l, b5_l, b6_l, b7_l, b8_l, b9_l, b10_l)

```

We can use one tailed to test the prediction that angles will be larger in winners than losers 
Testing with a priori expectation it will be larger in winners than losers 

```{r}
wilcox.test(w, l, paired = TRUE, alternative = "greater")
```

## Test 4 Effect size
```{r}
# combine into a data frame first
df <- data.frame(
  value     = c(w, l),
  condition = rep(c("win", "loss"), each = 10),
  bird      = rep(1:10, 2)
)
wilcox_effsize(df, value ~ condition, paired = TRUE)
```